---
title: "Session 10 in class actitivy"
author: "Chenxin Li"
date: "2026-06-18"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, you will learn how to use the `patchwork` package to make composite figures. 
You will need to install `patchwork` using the package panel. 

We will go over: 

* Simple layout 
* Complex layout with `design` syntax 
* Control relative widths and heights 

# Packages
```{r}
library(tidyverse)
library(patchwork)
library(RColorBrewer)
```

# Simple layout
For the purpose of this lesson, let's load some pre-made figures to assemble.   
The data are taken from [Tenorio Berrío et al., 2025](nph.onlinelibrary.wiley.com/doi/10.1111/nph.20446)

Go to [Zenodo record 22047472](https://zenodo.org/records/22047472) 
to download "S10_exampleA.Rds" and "S10_exampleB.Rds" and copy them to the Data folder. 
```{r}
panelA <- readRDS("../Data/S10_exampleA.Rds")
panelB <- readRDS("../Data/S10_exampleB.Rds")
panelCA <- readRDS("../Data/S10_exampleCA.Rds")
panelCB <- readRDS("../Data/S10_exampleCB.Rds")
panelCC <- readRDS("../Data/S10_exampleCC.Rds")
```

Free feel to view each panel by calling their name.   
```{r}
panelA
panelB
panelCA
panelCB
panelCC
```
Let's say I want to have `panelA` on top of `panelB`, what do you do? 
To make an assembly, we need to use `wrap_plot()`, which is part of `patchwork()`.  
We need to specify the panels and how many rows and columns to assemble them into. 
In this case we just want `panelA` on top of `panelB`, so it will be 1 column and 2 rows. 
```{r}
wrap_plots(panelA, panelB,
           nrow = 2, ncol = 1)
```
If we want to add the letter "A" to the top left of `panelA`, 
and letter "B" to the top left of `panelB`, we will use the `tag` option in `labs()`. 
Then we save the new assembly as an R object `AB`. 
```{r}
AB <- wrap_plots(panelA + labs(tag = "A"),
           panelB + labs(tag = "B"),
           nrow = 2, ncol = 1)

AB
```
This is a representation of single cell RNA-seq data. 
Each dot is a cell. Cells are grouped and clustered by their transcriptome. 
Cell type identities are color coded. 
Panel B is a subset of cells from the experiment. 
Sub-populations of vasculature cell types were identified.  

# Controlling complex layout and relative sizes 
For complex layouts, we need the `design` syntax. 
`patchwork` has an intuitive way to arrange plots using the `design = ` argument. 
`design` is a character string with rows and columns. 

The `design` argument is a character string of uppercase letters (A, B, C, and so on).
Within this vector, we use new lines to indicate different rows in the composite.

If the `design` has two rows, the composite will have two rows. For example, 
```
"AB
 CD"
```

It specified that "part A" will be at the top left, "part B" will be at the top right, 
"part C" will be at the lower left, and "part D" will be at the lower right.  

If you need a panel to use the full row space or full column space, 
repeat the letter across the rows and columns. 

```
"AB
 CC"
```
For example, this design will have panel C using up the entire 2nd row of the assembly. 

```
"AB
 AC"
```
For example, this design will have panel A using up the entire 1st column of the assembly. 

Now let's do a real example. 
Let's say I want panel C to look like this, which itself is an assembly of CA, CB, and CC.  

| Column 1    | Column 2  |
| :---------: | :-------: |
| CA          | CB        |
| Blank       | CC        |


In this case, the design looks like this: 

```
"AB
 #C"
```

Let's say we also want the first column to be only 5% of the width of the second column, 
and the second row to be 2% of the height of the first row.  
You can specify relative heights and widths using `heights` and `widths` options in `wrap_plot()`. 

```{r}
wrap_plots(panelCA, panelCB, panelCC, 
           design = "AB
                     #C" , 
           widths = c(0.02, 1),
           heights = c(1, 0.02), guides = "collect")
```
Using `widths = c(0.05, 1)`, I specified the first column is 5% width of the second column.  
Using `heights = c(1, 0.02)`, I specified the second row is 2% height of the first row.  
Lastly, I used `guides = "collect"` to collect the legends to the right. 

Let's add the letter "C" to the upper left of this assembly and save the assembly as an object. 
We will only need to add the letter to panelCA. 

```{r}
C <- wrap_plots(panelCA + labs(tag = "C"),
                panelCB, panelCC, 
           design = "AB
                     #C" , 
           widths = c(0.02, 1),
           heights = c(1, 0.02), guides = "collect")

C
```
1. This is a heat map showing the expression of aliphatic glucosinolate biosynthetic genes 
across different vasculature cell populations. 
Each row is a gene, and each column is a cell population. 
The darker the color, the higher the expression. 
The larger the dots, the more cells are expressing the mRNA in each cell population. 
You can see the pathway is most highly expressed in subcluster 5. 

2. The pathway is divided into 3 stages: chain elongation, core formation, and side chain. 
The color bar on the right annotates which stage the biosynthetic gene belongs to.

3. Finally, the color bar at the bottom annotates which cell type each cell population (subcluster) belongs to.  

## Hierarchical assembly
Now we can assemble the assemblies and export it to our disk. 
```{r}
wrap_plots(AB, C, nrow = 1, ncol = 2, widths = c(0.8, 1))

ggsave("../Results/S10_example.png", width = 7.5, height = 6)
```
It looks nice! And it's publication ready. 

## Applying theme setting across panels 
One final note is that you can apply `theme()` settings across the assembly using the `&` syntax. 
Let's say we want the entire figure to have bigger text, like font size 12. 
The function is `theme(text = element_text(size = 12))` . 
We simply add that to the `wrap_plots()` line using `&`. 
```{r}
wrap_plots(AB, C, nrow = 1, ncol = 2, widths = c(0.8, 1)) &
  theme(text = element_text(size = 14))
```
# In-class activity 
## Q1
Use your own data from your research to generate a composite figure. 
Save it using `ggsave()`. Produce both a .svg copy and a .png copy.  

For those of you who do not have your own data, feel free to use publicly available datasets.   
Also feel free to explore the income (S6_activity1.csv), continent (S7_example1.csv), mortality (S7_activity1.csv), and birth rate (S7_activity2.csv) datasets to assemble a figure that conveys a coherent message.  


