---
title: "Session 1 in-class activity"
author: "Chenxin Li"
date: "2026-06-15"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This is a header 
## This is a sub-header 

If you change the output to html_notebook, 
it will automatically render an HTML file on save.   
You should be able to find the HTML file in the same folder as your .Rmd script. 

You can write any plain text you want here.   
You can _italicize_ or __bold__ text if you want.  
You can ~~cross out~~ text as well.  
You can start a new line using two spaces at the end of a line.  

You can start a list like this: 

* First item of list. 
* Second item of list. 
* Remember to leave an empty line before the first bullet point. 

Or like this: 

1. First item.
2. Second item. 
3. If you want numbered list. 

This RMarkdown is for you to familiarize with RStudio, 
and learn how to load packages into your session.   
At the start of your RMarkdown, you want to have a brief 
introduction section to describe what this script or notebook is doing. 

# Packages 
After the introduction section, you will start a new section to load your packages. 
To do that, you will need to insert a code chunk. 
The short cut is `ctrl-alt-i` on PC or `cmd-opt-i` on Mac. 
Now you have a code chunk.  
```{r}
# This is where R will run your code. 
# If you need to write comments here, you will need to start the line with a # sign. 
```

To load a package, use `library()` function and specify the package name 
inside the parenthesis.   
To run a code chunk, click the green triangle at the upper right of the code chunk.  
Or you can do `ctrl-enter` on a PC or `cmd-enter` on a Mac to run line-by-line. 

```{r}
library(tidyverse) 
```

You might get some warning messages regarding the version of R.  
As long as there are no error messages, you are all good. 
 
The tidyverse is a collection of packages for data science and visualization.  
These packages use a similar style of syntax and design principal.  
While it is not the fastest or keystroke efficient, we will adopt these packages 
due to how intuitive they are. 

# How to get help 
Let's say you want to know how to use the `head()` function. 
You type ?head in the console.  
RStudio will pull up the help page in the lower right corner.  
It should say "returns the first or last parts of objects" or something similar. 

Let's try that. The tidyverse comes with a few built-in datasets.   
One of them is called `storms`.   
Let's try using `head()` on `storms`.  

```{r}
head(storms)
```
 
By default,  `head()` prints the first 6 rows.   
You can print the first 10 rows by adding `head(n = 10)`, for example. 

```{r}
head(storms, n = 10)
```

# Checking dataset dimensions. 
You can check how big the dataset is by checking how many rows and columns there are. 
You will use the `dim()` function. 
Type ?dim in the console to learn how to use it. 

```{r}
dim(storms)
```

`dim()` returns 2 numbers.  
The first number is rows, and the second number is columns.  
This dataset has 19,537 rows and 13 columns. A big dataset! 

# In-class activity 
Start a new Rmd file. 
Change output from html_document to html_notebook.

# Q1 
Briefly introduce yourself. 
	* Which degree program are you in? 
	* What year are you in? 
	* If known, which lab are you part of? Very briefly describe what are you working on. 
	* What do you want to learn in 301? 

Feel free to use formatting options such as italic, bold, lists, and headers.  

# Q2
Load the tidyverse package. 

# Q3 
View the first 10 rows of the `starwars` dataset.

# Q4
How many rows and columns does `starwars` have? 

# Q5
Click Preview and Knit to HTML. Check if the HTML has been rendered properly. Upload to D2L. 
 

