---
title: "Session 2 in-class activity"
author: "Chenxin Li"
date: "2026-06-15"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this lesson, we will go over some R coding fundamentals.  
We will cover: 

1. Arithmetic
2. Data types in R 
3. Vectors 
4. Matrices 

# Let's do some math!
Adding, subtracting, multiplying, and dividing. 
```{r}
1 + 1
```
If you type 1 + 1 and run the line of code,  
you should get [1] 2. The `[1]` in the output means it's a single value. 
R understands the sequence of operation when you use parentheses. 
```{r}
1 + 1 * 5 / 2
((1 + 1) * 5)/2
```

## Mathematically functions. 
R has implemented many commonly used functions. 
Let's try square root first. It's `sqrt()`.  
```{r}
sqrt(2)
```
R has a log() function. Type ?log in the console to learn how to use it.  
You will see that the default is natural log.  
If you want to use log base 10, you will either do `log10()` or `log(base = 10)` 
```{r}
log10(1000)
```

## Scientific notations
You can use scientific notations in R using `e`.
For example, 10^6 is 1e6 and 1 ng is 1e-9 gram. 

```{r}
1e6 
1e-9
```

## Assigning variables and storing them in the memory. 
Let's say you have a variable x that is equal to 100. 
```{r}
x <- 100 
```

You will need to use the left arrow.  
Now if you call x, R will print its value. 
```{r}
x
```
Now if you check the upper right corner of RStudio in the Environment area, you should see x.  
Another way to check what variables you have stored in the memory is the `ls()` function.

```{r}
ls()
```
```{r}
x <- 200 
```

Using the same variable name will overwrite the variable.  
These variables are stored in the memory of R. After you close R, they will be gone.  
We will talk about how to write things into the hard drive in later lectures.  

# Data types in R 
R has a few basic data types: 

1. numeric 
  - integer 
  - decimal 
2. logical 
3. character 

## Numeric 
Numeric data should be self-explanatory. 

## Logical 
Logical or Boolean data is either TRUE or FALSE.  
Use all uppercase letters to denote them.  
Alternatively, they are denoted by single uppercase letters T and F. 
```{r}
TRUE 
FALSE
T
F
```
You will see that they are rendered as blue text in the code chunk. 

## Character 
One important thing to remember in R is that characters must be in quotes. 
```{r}
"x"
x
```

R will understand `"x"` as the character "x".  
R will understand `x` as the variable we stored earlier. 

```{r}
'x'
```

Single quotation marks also work, as long as they come in a pair.   

# Vectors 
Imagine a column in a data table. That's a vector in R. 
To create a new vector, you use the `c()` function. 
```{r}
my_vector1 <- c(1,2,3)
my_vector1
```
Note that all members of a vector must be the same type of data.  
E.g., all numeric, all logical, or all character, but not mixed. 
```{r}
my_vector2 <- c("a", "b", "c")
```

You can create vectors using the `seq()` function.  
Type ?seq in the console to learn how to use it. 
Say you want a sequence from 0 to 100, by increments of 10.   
```{r}
my_vector3 <- seq(from = 0, to = 100, by = 10)
my_vector3
```

You can also repeat a sequence many times using `rep()`. 
Let's repeat `my_vector1` 3 times.  
```{r}
my_vector1_repeated3 <- rep(my_vector1, times = 3)
my_vector1_repeated3
```

You can check the length of a vector using `length()`
```{r}
length(my_vector1_repeated3)
```

## Pull out specific members of a vector using indexes. 
Say we want the first, second, and eighth member of `my_vector1_repeated3`. 
```{r}
my_vector1_repeated3[c(1,2,8)]
```
## Arithmetic across vector 
When an arithmetic operation is applied to a vector,  
R will apply the operation to every element of the vector.  
```{r}
my_vector1_repeated3+1
```

```{r}
sqrt(my_vector1_repeated3)
```
## Summary stats across vector 
You can calculate `sum()`, average `mean()`, `median()`, and standard deviation `sd()`  
using built-in functions. For example: 
```{r}
sum(my_vector1_repeated3)
mean(my_vector1_repeated3)
```

# Matrix 
If you need to do linear algebra,  
you will need to use the matrix utility in R.  
The best way to create a matrix is to bind different vectors together.  

We have two ways to bind vectors together:

* by row: `rbind()`
* by column: `cbind()`. 

Example: 
```{r}
my_mat1 <- rbind(c(1, 2, 3, 4),
                 c(5, 6, 7, 8),
                 c(0, 0 ,0 ,0))

my_mat1
```
Also note that all members of the matrix need to be same data type. 

Like vectors, arithmetic and summary statistics work on matrix too. 
```{r}
sum(my_mat1)
mean(my_mat1)
```

You can check the dimension of matrix using `dim()`. 
```{r}
dim(my_mat1)
```

Like vectors, you can select specific rows and columns using indexes (square brackets[]). 
Say you want the first row: 

```{r}
my_mat1[1, ]
```
Or the 2nd column
```{r}
my_mat1[,2]
```
Because the matrix has 2 dimensions, the index now takes 2 numbers.  
The first number is the row, and the second number is the column. 
If you need everything in a row, you leave the second index blank.  
If you need everything in a column, you leave the first index blank. 

You can select multiple rows and columns.
```{r}
my_mat1[c(1,2),c(2,3)]
```

# Pipes  
You can build pipelines in R using the pipe `|>` operator. 
There are two pipe operators: `|>` or `%>%`.  
They do the same thing. 
`|>` is available through base R without loading any packages.  
`%>%` is only available if you load `library(tidyverse)`. 

How the pipe works is that it takes the output of the first function, 
and feed it as the input as the next function. 

Say you want to take `sqrt()`,   
then you take `log10()`, 
then you take the median.   

```{r}
my_mat1 |> 
  sqrt() |>
  log10() |>
  median() 
```

# In class activity 
## Q1
Using R, how would you calculate the square root of half of the average of the numbers 5.2, 10, 16.44, 17.3, and 10.6?

## Q2 
Create an object (variable) that stores the value 3^2 x 4^(1/8). Then overwrite that value with itself divided by 2.95. Print the result. 

## Q3
Find the formula online for converting degrees Fahrenheit to Celsius. Use a vector to convert the temperatures -40, 77, 20, 101, 120, and 212 in degrees F to C.

1.  Insert a new code chunk
2.  Make this matrix by binding 4 vectors together: 

    | 1 | 1 | 2 | 2 |
    | 2 | 2 | 1 | 2 |
    | 2 | 3 | 3 | 4 |
    | 1 | 2 | 3 | 4 |

  and save it as an item called `my_mat2`.

3. Select the 1st and 3rd rows and the 1st, 2nd and 4th columns, and save it as an item.

4. Take the square root for each member of my_mat2 and then take log2(). 
Find the maximum value afterwards. Use the pipe syntax. The command for maximum is `max()`.
