---
title: "Session 4 in-class activity"
author: "Chenxin Li"
date: "2026-06-16"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this lesson, we will go over:

1. Data structures
2. Boolean operations 
3. Strings  

# Packages 
We will need the tidyverse for strings.  
Let's just load it now and not worry about it later.   
```{r}
library(tidyverse)
```

# Data structures 
The 3 data structures we will go over today are:

1. Factors
2. List 
3. Data frame 

## Factors
Factors are vectors with defined levels. 
Factors are treated as discrete variables in R. 

For example, the months of the year can be understood as a factor.  
Let's say we are doing an analysis in which it's important for R to understand  
that January comes before February and so on. 
R has built-in month names and abbreviations. 
```{r}
month.abb
```

To specify the order, we will need to define it using the `factor()` function.  
Inside the `factor()` function, we will need to use `levels = ` option to  
define the intended order of this factor. 

```{r}
months_f <- factor(month.abb, levels = c(
   "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
   "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
))

months_f
```
Make sure you type in the levels in the intended order. 
Now if you print `months_f`, the output contains the Levels line:  
Levels: Jan Feb Mar Apr May Jun Jul Aug Sep Oct Nov Dec , which is the right order. 

You can reorder a factor by changing the order with the `levels = ` option.  
Let's say we are planting corn, for which we plant in April and harvest in October.  
So we need to set April as the first month, and October as the last month. 

```{r}
months_corn <- month.abb[4:10] # Use index to call out the 4th to 10th month. 

months_corn_f <- factor(months_corn, levels = c(
    "Apr", "May", "Jun", 
   "Jul", "Aug", "Sep", "Oct"
))

months_corn_f
```

## List 
List is a convenient way to collect data. 
It has no restrictions on what types of data can be part of the list.  
E.g., you can put a vector, factor, data frame, character, all into a list.  
You can use list as a way to organize a project.  
E.g., you can put all the objects associated with a project under a list. 

Here is a toy example: 
```{r}
project1 <- list("A", c(1,2), months_corn_f)
```

You can pull out individual members of the list using index.  
But in this case, you use double square brackets. 

```{r}
project1[[1]]
```

## Data frame 
Data frame is a data table. It has rows and columns. 
We will use the `storms` dataset from tidyverse as an example. 
```{r}
head(storms)
```

Note that all rows in a given column must be the same data type. 
For example, the `name` column is all character.  
The `year` column is all numeric.  
The `status` column is a factor.  
You can check the data type for each column by using `head()`. 
The data type should be shown under the column name. 

You will notice some data frames have column names. 
You can pull out columns by their name using the `$` syntax. 
Say I want to pull out the name column. I would do `storms$name`
 
The easiest way to create a data frame is prepare it in Excel and load it in R.  
We will go over how to do that in the next lecture. 
But if you need to create a small data frame, you can do that in R as well.  

Say I have 3 apples, 4 bananas, and 2 pears. 
I want to organize this into a table and call this table fruits.
What do I do?

```{r}
fruits <- data.frame(
  name = c("apple", "banana", "pears"),
  count = c(3, 4, 2)
)

fruits
```
To make a new table, you can use the `data.frame()` command.
Inside `data.frame()`, you then type out each columns as vectors.
In this example, the columns are name and count.

# Boolean operations
Also known as logical operations. 

Here are a few things to remember: 

* `==` means "equal to". `A == B` means "object A is equal to B". 
* `=` means assign value. `A = B` means "let value of A be B". 
* `<`, `<=`, `>`, `>=` are less than, less and equal to, greater than, and greater and equal to, respectively. 

Logical operations work on both single values and vectors. 
```{r}
A <- c(1, 2)
B <- c(1, 2) 
C <- c(1, 3)

A == B
A == C
```
When you call a Boolean operation, it applies the operation to each member of the vector.  

There is another set of logical operations. 

* `&` means "and" 
* `|` means "or" 
* `!` means "not" 
* `%in%` means belong to or is a subset of. 

Here is how to use them. 

`A & B` means both A and B must be TRUE.  
`A | B` means either A or B or both is TRUE. 
`A != B` means A is not equal to B. 
`A %in% X` means A is a subset of a set X. 

Let's say we want to know if apple is part of the fruits we have in the fruits data frame. 
```{r}
A <- "apple"
A %in% fruits$name
```
In this case it's TRUE. If you check `fruits$name`,  
it should say `c("apple", "banana", "pears")`,  
and therefore "apple" is part of it.  

# Strings 
Any string of symbol, such as “Hello World!”, including spaces, numbers, and symbols. 
```{r}
"Hello World!"
```
You can check the number of characters in strings using `nchar()`.
```{r}
nchar("Hello World!")
```
In this case it has 12 characters including space and symbols. 

You can paste any string together using `paste()`.  
You can preview any string using `cat()`. 

In addition, you can start a new line or insert white space: 

* New line: "\n" (does not count towards characters)
* Tab: "\t" (does count towards characters) 

Let's say you did a t-test, 
where the p value is 0.04.  
And you want to print that in the code. 

```{r}
p <- 0.04
test <- "t-test" 

message <- paste("p = ", p, "\n", test, sep = "" )
cat(message)
```
By default, `paste()` inserts spaces (`sep = " "`) between strings. 
You can bypass that by changing the `sep` option. 

## Regular expression. 
Regular expression (regex) is a tool to match patterns in string. 
R has implemented regex. 
The stringr package (part of tidyverse) provides a lot of utility for strings and regex, which is what we will use.  

Here are some useful tools (as implemented in stringr):  

* Start of line: `"^"`
* End of line: `"$"` 
* Any one character, including space, number, and symbols: `"."` 
* Any one numbers: `"\\d"`
* Multiple characters: `"[]"`  
* All letters: `"[:alpha:]"` 
* Lower case letter: `"[:lower:]"`
* Upper case letters: `"[:upper:]"`
* Letters, numbers, and symbols: `"[:graph:]"`

If you need one or more of the same pattern, add `"+"` to the end of the pattern. Example: 

* One or more numbers: `"\\d+"`
* One or more lower case letters: `"[:lower:]+"`
* One or more upper case letters: `"[:upper:]+"`

And so on. Note that `".+"` will match anything,   
as it stands for "one or more of any character, including space, number, and symbols".

If you need to detect a symbol that is already used to mean something else, 
you add `"\\"` before it. For example, 

* The symbol `"."` is `"\\."`
* The symbol `"^"` is `"\\^"`
* The symbol `"$"` is `"\\$"` 

And so on. 

There are a couple of exceptions to this: 

* The forward slash is just `"\\"` rather than `"\\\"`. 
* The quotation mark `"` is just `"\""` rather than `"\\""`.

But these are rare.    

Do not memorize. 
You can google or refer to this lesson if you need a refresher. 

## Using regex for pattern detection, extraction, and replacement with stringr functions. 
Let's use a text string as an example: 

```{r}
test <- "abcABC123\t.!?\\(){}\n"
cat(test)
```
This is the test string. 

Stringr functions all start with 'str_' follow by a verb. 

* `str_detect()` for detection. 
* `str_extract()` for extraction.
* `str_replace()` for replacement.  

### Detection 
Let's say we want to detect the string "abc" in our test string. 
```{r}
str_detect(test, "abc")
```
In this case it returns TRUE, as abc is part of the string. 

If you want to detect "Abc", it will be
```{r}
str_detect(test, "Abc")
```
In this case, it will be FALSE, as the exact string upper case A followed by lower case bc is not present in the string. 

If you want to detect upper case A or lower case b or lower case c,  
you will need the square brackets syntax. 
```{r}
str_detect(test, "[Abc]")
```
In this case it is TRUE.  

### Extraction
Let's say we want to extract part of the string containing a, A, or 1.  
```{r}
str_extract(test, "[aA1]")
```
You will see it only returns "a", even though A and 1 are also present in the test string.   
This is because the default behavior of `str_extract()` stops at the 1st occurrence. 
What you will need is `str_extract_all()`. 

```{r}
str_extract_all(test, "[aA1]")
```
If you do that, you should get a list containing 3 elements: "a", "A", "1". 

Let's say you want to extract all numbers in your string: 
```{r}
str_extract_all(test, "\\d")
```

### Replacement 
Replacement is very much like extraction,  
but now we need to specify the string to replace with. 

Let's replace abc to upper case letter.  
Remember, if you need to detect all occurrences of a pattern,  
don't forget to use `str_replace_all()`. 
```{r}
str_replace_all(test, "abc", "ABC")
```
If you want to change from upper to lower case letters, 
or vice versa, simply use `str_to_upper()` or `str_to_lower()`. 

```{r}
str_to_upper(test)
```
Say you want to remove all numbers,  
what you need to do is replace all numbers with "", which is quotation marks with nothing inside.  
```{r}
str_replace_all(test, "\\d+", "")
```
Now all the numbers are gone. 

# In class activity 
## Q1 
The academic year at MSU starts in Aug and ends in May. 
Create a factor starting in Aug and ends in May. 

## Q2 
1.  Use the following info to make a data frame and save it as an item called "grade".
    Adel got 85 on the exam, Bren got 83, and Cecil got 93.
    Their letter grades are B, B, and A, respectively.
    (Hint: How many columns do you need to have?)

2. Pull out the column with the scores.
    Use the `$` syntax.
    
## Q3 
Write a Boolean operation to evaluate if square root of 5 is greater than pi.

## Q4
Here is a sequence of a protein called [CPC from Arabidopsis](https://www.uniprot.org/uniprotkb/O22059/entry): 

`MFRSDKAEKMDKRRRRQSKAKASCSEEVSSIEWEAVKMSEEEEDLISRMYKLVGDRWELIAGRIPGRTPEEIERYWLMKHGVVFANRRRDFFRK` 

Here is the sequence of a bHLH-binding motif: 
`[D/E]Lxx[K/R]xxxLxxxxxxLxxxR`, where x is any amino acid residues,  
[D/E] means D or E, and [K/R] means K or R. 

1. Does this protein contain a bHLH binding motif? 
Print its sequence. 
Hint: what should the regex be? Make sure your protein is read as a single line.

2. Mutate all leucine (L) within the bHLH binding motif to lysine (K). 
Print the full length protein sequence of the mutant. 


## Q5
You have 3 samples:
`c("sample-1", "Sample-2", "sample_3")`

Use `str_replace_all()`to change the upper case "S" to lower case "s" and  
change all hyphens to underscores. 
Hint: how many steps do you need? 

## Q6 
Run `str_extract_all(test, "\\d")` and `str_extract_all(test, "\\d+")`. 
Can you explain what causes the differences in the results? 








