---
title: "Session 5 in-class activity"
author: "Chenxin Li"
date: "2026-06-16"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This lesson we will go over:

1. File input/output (I/O) in R. 
2. How to make an R script. 
3. How to write custom functions.
4. Other control structures.
5. Applying a function across a list or vector. 

# Packages
We will need the tidyverse as well as a new package called `readxl`. 
`readxl` is a package that allows you to directly read an Excel spreadsheet to R. 
Use the Package panel at the lower right corner of RStudio to install `readxl`. 

```{r}
library(tidyverse)
library(readxl)
```


# File I/O. 
R works in the memory, not the disk (hard drive).  
The hard drive or disk is the permanent storage of your data. 
Your data is stored even if your device is powered off.  
In contrast, the memory is only short-term storage.  
The data is stored temporally for applications to use and modify.  
However, the data will not be saved when the app is closed,  
or when the device is powered off. 

This means we need to read data into R (file input), 
and after we are done,  
we need to write the results onto the disk (file output). 

## Specifying the path and file 
Let's say we want to read an Excel file, we need to use `read_excel()`. 
Inside `read_excel()`, we have to specify where the file is. 
In this example, the file is at "/Users/cxli9/Desktop/Li/MSU/Teaching/Bioinformatics/301-F2026/Data". 
The location will be different for your computer. 
Note that R uses the forward slash "/" for denoting folders and sub-folders.  

"/Users/cxli9/Desktop/Li/MSU/Teaching/Bioinformatics/301-F2026/Data" is the absolute path. 
You can also navigate your computer in R using the relative path. 

By default, R is reading into the folder where your current R/Rmd script is at. 
If you type "..", it will go one folder up. 
If you press `Tab`, it should show what other folders or files are in the current directory. 

```{r}
S5_example1 <- read_excel("../Data/S5_example1.xlsx")
head(S5_example1)
```

This is a tracking sheet Li used to track his job searches in 2024. 
Each row is a position that he applied to.  
Each column is a different attribute of the positions. 

## Reading tables of different formats. 
In addition to Excel, some of the most common file formats are: 

* .csv (comma separated values) 
* .txt (plain text, values separated by tab) 

In addition, you want to check if the data file comes with headers (also known as column names). 

### Read csv
we will use `readr` functions (part of tidyverse) to read csv or plain text files. 
Let's read a data table called "S5_example2.csv" in the Data folder. 
```{r}
S5_example2 <- read_csv("../Data/S5_example2.csv",
                                  col_names = T)
head(S5_example2)
```

In this case the data file comes with its own column names.  
By default, `readr` functions has `col_names` option set to TRUE. 
You can print the first 6 lines of the table using `head()` to check if the table was parsed correctly. 

### Read plain text. 
To read plain text, you will use `read_delim()`.  
In this case you will also need to specify if values are separated by tab or white space `" "`. 
If you are not sure, try both and see if the table was parsed correctly. 
Let's read a data table called "S5_example3.gtf" in the Data folder.
```{r}
S5_example3 <- read_delim("../Data/S5_example3.gtf", col_names = F, delim = "\t")
head(S5_example3)
```

In this example, the data file has no column names, so we set `col_names` to FALSE.  
Also, this data file uses tab `"\t"` to separate values.  

## File output. 
Writing files to the disk is the reverse operation of reading files. 
In this case, you need to specify which object in the environment to write, and the path.  
Make sure the target directory actually exists.  
You will use functions such as `write_delim()`(for plain text) and `write_excel_csv()` for csv.  
You can also reformat the data files if you want. 
For example, now I want to store `S5_example1` as a plain text file rather than Excel. 
Let's save it to the Results folder. 

```{r}
write_delim(S5_example1, "../Results/S5_example1.txt", 
            delim = "\t", col_names = T)
```

You need to specify values to be separated by tab (`"\t"`) and that the file has column names. 

## Directly reading/writing R data objects. 
You can also directly read an R object from the disk or write an R object to the disk.  
This could be helpful if you just want to permanently store an intermediate result from R.  

Let's say you have a result that you want to store. 
```{r}
important_result <- list(result1 = c("A", "B", "C", "X", "Y", "Z"),
                         result2 = c(1, 2, 3))
important_result
```

You can store it on disk with `saveRDS()`. Let's save it to the Results folder.
```{r}
saveRDS(important_result, "../Results/S5_important_result.Rds")
```
Don't forget to specify the path and the file name. 

You can retrieve it using `readRDS()`.
```{r}
important_result <- readRDS("../Results/S5_important_result.Rds")
important_result
```

# R script. 
An R script is plain text that R can read and execute. 
Every line of text is interpreted as code. Non-code text must start with # at the start of line. 

Good practice: write down title, usage, date created, author, last modified, input, output, packages required. 
Check "S5_Example_script.R" in the Scripts folder for an example. 

# Functions 
You can write custom functions in R. 
For example, the function logit is not implemented in R. 
logit is defined as for a number p between 0 and 1, 
logit = log(p/(1-p)).  

The syntax for function is `function(parameters){what to do}`. 
You will need to use the curly brackets `{}`. 

```{r}
logit <- function(p){
  log(p / (1-p) )
}
```

Once defined, the custom functions can be called like any built-in functions. 
```{r}
logit(0.5)
```
# Other control structures 
R can perform 

* If statement
* for loops
* while loops 

like other programming languages. There are better ways to do these in R.   
Here are some examples. 

## if statement 
The syntax for if statement is 
`if (condition) {do something} else {do something else}` 

```{r}
x <- 1
y <- 2

if (x > y) {
    paste(x, ">", y)
  } else{
    paste(x, "<=", y)
  }

```
## for loops
You use for loops to apply a function across a list or vector. 

The syntax for for loops is 
`for(list or vector){do something}` 

```{r}
some_vector <- c("A", "B" , "C", "D", "E")

for(index in 1:length(some_vector)){
		cat(index, some_vector[index], "\n")
} 
```

You can nest for loops in another for loop.  
I _do not_ recommend. 


## While loops
Syntax for while loops is  
`while(condition){do something}`

```{r}
index <- 5 
	while(index <= 10){
	cat(index, "\n")
	index <- index + 1
}

```
Be super careful when you do while loops! You can get stuck in an infinite loop. 

# Applying a function over a list. 
R has a set of `apply()` functions, which is a better way to do for loops.  
If you want to apply a function across a list, you use `lapply()`.  
If you want to apply a function across a vector, you use `vapply()` and so on.  

The syntax is `l/vapply(list/vector, function, additional parameters)`.  

Let's say I have a bunch of files from an analysis. I need to read them all in R. 
The results files are in "../Data/batch_import_example1" folder. 
First you will create a list for each files in the folder. 
They are all coverage files in .cov format. 

```{r}
file_list1 <- list.files("../Data/S5-batch_import_example1/", pattern = ".cov",
                         full.names = T)
file_list1
length(file_list1)
```
`full.names = T` option preserved the path, which is important for `read_delim()`. 
Looks like we have 21 files. 
Then we apply `read_delim` for all of them. 

```{r message=F}
result1 <- lapply(file_list1, FUN = read_delim, 
      delim = "\t", col_names = F)
```
We are applying `read_delim()` to every file name in the `file_list1`.   
Note that when you are looping a function, you do not add parentheses `()` after the function name  
because we are simply referring to the function name here.  
Also note that `read_delim()` also requires `col_name` and `delim` parameters. 
There additional parameters are specified in `lapply()` instead. 
R understands that these parameters are applied to everything in the list.  

```{r}
table1 <- result1[[1]]

head(table1)
```

Now you can see that result1 is a list of 21 tables. 
You can call out each table using index, such as `[[1]]`.   
We will go over how to convert a list of data frames into a single data frame in later lectures.   

# In class activity 
## Q1
Load S5_activity1.xlsx in the Data folder into R.  
Hint: this is an Excel spreadsheet. 

## Q2
Load S5_activity2.csv in the Data folder into R. 
Then write it as a plain text file in the Results folder. 

## Q3
Batch import data files from the `Data/S5-batch_import_example2`. 
Then save the list of data tables as an Rds object in the Results folder.  

## Q4 
The logistic function has the formula: y = 1/(1+exp(-x)). 
Write an R function for the logistic function. 
What is the value of the logistic function at x = 0? 
What is the value of the logistic function at x = 1000?
What is the value of the logistic function at x = -1000? 

 

 



