---
title: "Session 6 in-class activity"
author: "Chenxin Li"
date: "2026-06-17"
output: html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, we will go over basic operations to tabular data. 
We will go over: 

* What is a tidy data frame? 
* Wide-to-long and long-to-wide operations. 
* Basic verbs for data frames. 
* Downstream processing of batch-imported tables. 

# Packages
For this lesson, we will need `readxl` to read an Excel file.  
In addition, we will need the `tidyverse`. 
```{r}
library(readxl)
library(tidyverse)
```

# What is tidy data?
A data frame is tidy if: 

* Each row is an observation. 
* Each column is a variable. 

Let's load an example.
We will use Li's job search tracking sheet (example 1 from Session 5). 
```{r}
S5_example1 <- read_excel("../Data/S5_example1.xlsx")
S5_example1
```

In this table, each row is a position that I applied to. 
It has 19 rows, so there are 19 observations.  
Each column is a variable (or attribute): school, department, URL, deadline, key words and so on. 
So this is a tidy data frame. 

# Wide-to-long operation 
Now let's look at another example. 
```{r}
S6_example1 <- read_csv("../Data/S6_example1.csv", col_names = T)
head(S6_example1)
```

In this example, we have crops (barely, canola, corn, and so on) as rows. 
Each column is an year. 
The value of the table is the yield at any given year. 
Is this a tidy data frame? 

```{r}
dim(S6_example1)
```
If we check the dimension of the data, we will find that there are 8 rows and 79 columns. 
So we have 8 crops and 78 years of data (not counting the 1st column because it's the crops themselves). 
In total, we could have 8*78 = 624 observations, so the tidy form of this data should have up to 624 rows. 

There are up to 78 years of data for each crop, but the observations are spread out in 78 columns.  
Observations spreading out in columns is sometimes called the wide format.   
The wide format is easy for data collection, but unfortunately it's not how R would like to read data.   
R likes tidy data, and tidy data frames need each observation to be in its individual row.   

To convert a wide table into a tidy table, the command is `pivot_longer()`.  
The syntax is `pivot_longer(columns, names_to = "name", values_to = "value")` 

What we need is 3 columns, corresponding to the 3 variables we have in hand. 

1. Crop
2. Year 
3. Yield 

We will use the pipe syntax here (`|>` or `%>%`). 
We start with our data `S6_example1` and feed it into `pivot_longer()`. 
In this example, we need to select all columns that are not `Crop`.  
The former column names will be stored as a new variable called "Year".  
The values of the old table will be stored as a new variable called "Yield".  
Let's give that a try. 

```{r}
S6_example1_long <- S6_example1 %>% 
  pivot_longer(cols = !Crop, names_to = "Year", values_to = "Yield") 

head(S6_example1_long)
```

```{r}
dim(S6_example1_long)
```
Now if you check the dimensions of the data,  
you will see it has exactly 8*78 = 624 rows and 3 columns, 
which is what we should expect if each row is an observation.  
However, there is just one problem.  
If you look at the first 6 rows of the data frame using `head()`, you will see that  
the `Year` column is recorded as a character (`<chr>` shown under the column name) instead of numeric variable.  
We will need to fix that later. 

Let's try another example. This time it's a hypothetical example that I made up.
```{r}
#ignore this chunk. It simulates the data.
example_data <- data.frame(
  "sample1" = c(1, 2, 3),
  "sample2" = c(2, 2, 2),
  "sample3" = c(3, 2, 1)
) %>% 
  mutate(gene = c("gene1", "gene2", "gene3"))

head(example_data)
```
In this experiment, 
you are looking at the expression of three genes (gene1 to gene3) in three samples (sample1 to sample3).
So you have 3 * 3 = 9 observations.
This is not a tidy data frame because the 9 measurements of expression are spread out in 3 columns.
We can tidy it using the `pivot_longer()` command:

```{r}
example_data_tidy <- example_data %>% 
  pivot_longer(names_to = "sample", values_to = "expression", cols = c(sample1, sample2, sample3)) 

example_data_tidy
```

1.  The "name" column is now "sample": it houses which original columns the data came from (sample1 to sample3).
2.  The "value" column is "expression": it is the variable that was actually measured.
3.  The columns to collect data from were sample1 to sample3. Alternatively, `cols = !gene` (all columns but gene) should also work. 

Now this is a tidy data frame. Each row is an observation. Each column is a variable.
We have 3 variables: sample (3 samples), genes (gene1 to gene3), and expression (numeric).

# Long-to-wide operation. 
Some statistical tools require the data to be in the wide form. 
The reverse operation is `pivot_wider()`. 
You will need to specify which column will become the new column names (`names_from`). 
And which column to fill the data table with (`values_from`).  

```{r}
example_data_wide <- example_data_tidy %>% 
  pivot_wider(names_from = "sample", values_from = "expression")

example_data_wide
```
We will see that this is the exact reverse operation to `pivot_longer()`. 

# Basic operations to a data frame.
The 3 most commonly used basic verbs for data frames are `select()`, `filter()`, and `mutate()`. 

## Selecting columns by name. 
Using our tidy example1 data now,
say we only want the columns Crop and Yield and leave out the rest of the columns.
The command is simple, just `select()` the columns you want.

```{r}
head(S6_example1_long)

S6_example1_long_2 <- S6_example1_long %>% 
  select(Crop, Yield) #just selecting Crop and Yield 

head(S6_example1_long_2)
```

Now we just get two columns.

### Filter by Boolean logic 
The more useful subsetting function is subsetting by rows based on the values of one or more columns.
The command is `filter()`.  

In the example, let's say we only want years that have data because data were not collected in certain years for some crops.  
```{r}
S6_example1_long_3 <- S6_example1_long %>% 
  filter(is.na(Yield) == FALSE)  # filter for rows that the Yield value is not NA  

dim(S6_example1_long)
dim(S6_example1_long_3)
```
In this example, we want to find Yield values that are not `NA`. 
R has a built-in function for that, which is `is.na()`.  
We want Yields that are not `NA`, so we will use the Boolean logic `is.na(Yield) == FALSE`.  
Remember, in R, you use `==` to mean "equal to". 


Let's say we only want corn. 
```{r}
S6_example1_long_corn <- S6_example1_long %>% 
  filter(Crop == "corn") 

head(S6_example1_long_corn)
```

* "corn" has to be in quotes since this is referring to characters in R. 
* In this case "corn" has a lowercase "c". We need exact matches, which is case sensitive. 

### Filter with complex conditions 
Let's say you want to filter for corn and wheat. 
```{r}
S6_example1_long_corn_wheat <- S6_example1_long %>% 
  filter(Crop == "corn" | 
           Crop == "wheat") 

head(S6_example1_long_corn_wheat)
```

Remember the vertical bar `|` means "or" in R. In this chunk you are filtering for either corn or wheat.
You shouldn't use `&` ("and") here, because there is no crop that is both corn and wheat.

An alternative approach is to use the `%in%` syntax here. 
```{r}
S6_example1_long_corn_wheat <- S6_example1_long %>% 
  filter(Crop %in% c("corn", "wheat")) 

head(S6_example1_long_corn_wheat)
```

### Filter with string and regex 
Another alternative approach is to use string detection.  

```{r}
S6_example1_long_corn_wheat <- S6_example1_long %>% 
  filter(str_detect(Crop, "orn|heat")) 

head(S6_example1_long_corn_wheat)
```

## Create new columns 
To create new columns you use `mutate()`.  
It is just as versatile and powerful as `filter()`.  

### Simple transformations 
```{r}
head(S6_example1_long)
```

In this example, the `Year` is stored as a character variable. Let's change that to a numeric variable. 

```{r}
S6_example1_long_4 <- S6_example1_long %>% 
  mutate(Year = as.numeric(Year)) 

head(S6_example1_long_4)
```

You should note that when you create a new column but you use an existing column name (in this case Year),  
the existing column will be overwritten by the new column. 

### Create new column by formula 
You can do any arithmetic operations to a numeric variable within `mutant()`.  
More importantly, you can specify a different formula based on values of other values.  

Let's say the yield for every crop other than cotton, canola, and rice were recorded in bushel/acre. Cotton, canola, and rice were recorded in lbs.  
The number of pounds in a bushel vary by grain.  
Let's say there are 60 lbs/bu for wheat, 56 lbs/bu for corn,
48 lbs/bu for barley, 60 lbs/bu for soybean, and 56 lbs for sorghum. 
Let's convert all the yield units to lbs/acre.  

To do that, we will need a new tool `case_when()`.  
`case_when()` is an extremely useful command within `mutate()`.
The syntax is `case_when(condition ~ outcome)`.
For example, when Crop is barley, we need to apply a value of 48 lbs/bu. 

```{r}
S6_example1_long_lbs <- S6_example1_long %>% 
  mutate(Yields_lbs = case_when(
    Crop == "barley" ~ Yield * 48,
    Crop == "corn" ~ Yield * 56, 
    Crop == "soybean" ~ Yield * 60,
    Crop == "sorghum" ~ Yield * 56, 
    Crop == "wheat" ~ Yield * 60,
    T ~ Yield
  ))

head(S6_example1_long_lbs)
```

The `T ~ ...` inside `case_when()` stands for "for the rest that are not the above". 
I hope you can appreciate that this is very powerful and convenient. 
The syntax is relatively easy to understand.  
This approach also saves you from having to write nested if statements. 

### Create a new column using values of other columns. 
Let's say you want to classify crops into different categories.  
We have 8 crops and we want to classify: 
 
* cotton as a fiber crop 
* canola and soybeans as an oil crop 
* rice, corn, wheat, barely and sorghum as grains 

```{r}
S6_example1_long_5 <- S6_example1_long %>% 
  mutate(Category = case_when(
    Crop == "cotton" ~ "fiber",
    str_detect(Crop, "canola|soy") ~ "oil",
    T ~ "grain"
  ))
```

# Downstream processing of batch-imported tables. 
Last lecture we talked about how to apply a function across a list to batch import data files into R. 
This lecture we will pick up from there and continue processing them. 
We will use the files in "../Data/batch_import_example1" folder. 
```{r}
file_list1 <- list.files("../Data/S5-batch_import_example1/", pattern = ".cov",
                         full.names = T)
file_list1
length(file_list1)
```
We have 21 files. 

Tidyverse has the utility to bind data frames as rows of each other if they have the same columns.  
To do that, you use the `bind_rows()` function. 
Inside `bind_rows()`, you have the `.id = "id"` option. 
This will create a new column called `id` to keep track of the original identity of the rows. 

```{r message=F}
result1 <- lapply(file_list1, FUN = read_delim, 
      delim = "\t", col_names = F) %>% 
  bind_rows(.id = "id")

head(result1)
```

In this case, the rows from the original file 1 will have an id value of 1. 
The rows from the original file 2 will be have an id value of 2, and so on. 

You can also append the original file name to this big result table by doing some quick data wrangling. 
We will talk about how to do that in the next lesson, which involves combining two or more data tables by column. 

# In class activity 
## Q1 
Read S6_activity1.csv from the Data folder into R. 
Data source: the [Gapminder foundation](https://www.gapminder.org/data/). 

1. Turn the data table into a tidy data frame. 
In this example, each row is a country, each column is a year, and the values are average person income in USD. 

2. Produce a data table with just data for Bangladesh and Sweden, in years 1945 (when WWII ended) and 2010. 

3. Let's say for countries with income between 1000 to 10,000 dollars per year, they are called "fed".
For countries with income above 10,000 dollars per year, they are called "wealthy".
Below 1000, they are called "poor".
Use this info to make a new column called "status".
Hint: you will have to use case_when() and the "&" logic somewhere in this chunk. 

## Q2
Batch import data files from `Data/S5-batch_import_example2`.  
Convert it to a single data frame with the original identity of the rows tracked with an id column.  
