---
title: "Session 7 in-class activity"
author: "Chenxin Li"
date: "2026-06-17"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, we will go over:

* How to join multiple data frames together by matching columns. 
* How to use the grouping function. 
* How to perform summary statistics. 

# Packages
For this lesson, we only need tidyverse. 
```{r}
library(tidyverse)
```

# Joining data by column. 
Oftentimes you'll need to add different columns from another data frame.
The jargon for that is joining. 

There are a few commands for joining in R:

1.  `left_join(X, Y)` keeps all the rows in X.
2.  `right_join(X, Y)` keeps all the rows in Y.
3.  `inner_join(X, Y)` only keeps rows that are in common.
4.  `full_join(X, Y)` keeps all rows, filling in `NA` when values are missing from either one.

Usually, `inner_join()` will work well. 
When you perform downstream analyses you might filter out `NA` anyway. 

Let's continue with where we left off last lesson with file batch import and processing. 
We will use the files in "../Data/batch_import_example1" folder. 
```{r}
file_list1 <- list.files("../Data/S5-batch_import_example1/", pattern = ".cov",
                         full.names = T)
file_list1
length(file_list1)
```
These were the same 21 files we've been working with. 

```{r message=F}
result1 <- lapply(file_list1, FUN = read_delim, 
      delim = "\t", col_names = F) %>% 
  bind_rows(.id = "id")

head(result1)
```
We have a single data frame with an additional column called id, 
which tracks the original identity of the rows. 


Now let's append the original file name to the big data frame. 
We just need to do some quick data wrangling. 
```{r}
my_og_file_names <- data.frame(
  file_names = file_list1, 
  id = as.character(1:length(file_list1))
)

my_og_file_names
```

We first created a data frame where one column is the file name, 
the other column is the id, matching what is in the big data frame.
I also went ahead and changed the id column from numeric to character 
because the id column was stored as character in the big data frame.  

Now we just need to join them. Here we can use `full_join()`  
because there are 21 files and the id ranges from 1:21, each matching a file name.   

```{r}
result1_with_file_names <- result1 %>% 
  full_join(my_og_file_names, by = "id")

head(result1_with_file_names)
```

You can also use string operations to remove extra text in the file name if you want. 
```{r}
result1_with_shorter_file_names <- result1_with_file_names %>% 
  mutate(file_names = str_remove_all(file_names, "../Data/S5-batch_import_example1/50KB.|_ssl.bed.cov"))

head(result1_with_shorter_file_names)
```

Let's try another example. 
We will load S6_activity1.csv into R, which is income data from different countries across years. 
The data were downloaded from the [Gapminder foundation](https://www.gapminder.org/data/).
```{r}
income <- read_csv("../Data/S6_activity1.csv") 
head(income)
```

You will see that it's not a tidy data frame yet so let's address that.
```{r}
income_tidy <- income %>% 
  pivot_longer(cols = !country, names_to = "year", values_to = "income")

head(income_tidy)
```

Now let's say we want to know which continent each country is on. 
To do that we will need another data table. 
I downloaded the data table from [Our World in Data](https://ourworldindata.org/grapher/continents-according-to-our-world-in-data?tab=table).   
I saved it as "S7_example1.csv" in the Data folder. 
```{r}
country_by_continent <- read_csv("../Data/S7_example1.csv")
head(country_by_continent)
```

We will check the first 6 rows first. 
The information we need is country and continent. 
The country information is stored as a column called "Entity".  
The continent information is stored as column called "World region according to OWID".   
A good practice is to not include space in column names, so let's address that. 
You can work around column names with space by using the pseudo-quote "``". 
```{r}
country_by_continent <- country_by_continent %>% 
  mutate(continent = `World region according to OWID`) 

country_by_continent
```

Now we just need to join them. 
You will notice that in the `income_tidy` table, countries are in the column name "country".  
But here countries are in the column "Entity".  
We can workaround that by either making a new column called country using `mutate()`,  
or use a bit of a weird syntax in joining. 
Let me show you how. 

```{r}
income_tidy_continent <- income_tidy %>% 
  inner_join(country_by_continent, by = c("country" = "Entity"))

head(income_tidy_continent)
```

Here, `by = c("country" = "Entity")` specified that we want to match the "country" column from table 1 to the "Entity" column from table 2. 


# Operations by group. 
Tidyverse has a `group_by()` function that does not modify the data frame but changes how subsequent operations are done.  
Let's look at this example. 

```{r}
# Ignore this chunk - just making some hypothetical data 
example_transcriptome <- data.frame(
  gene = c("gene1", "gene2", "gene3"), 
  timepoint1 = c(1, 100, 100),
  timepoint2 = c(5, 500, 50),
  timepoint3 = c(10, 1000, 10)
)

example_transcriptome
```
Let's say you are given a table of gene expression. 
Each row is a gene, and each column is sample, in this case 3 time points. 
The values are the normalized expression value in units of transcripts per million (tpm). 

We can make it into a tidy data frame first. 
```{r}
example_transcriptome_long <- example_transcriptome %>% 
  pivot_longer(cols = !gene, names_to = "sample", values_to = "tpm") 

example_transcriptome_long
```

## Calculate z-score from a tidy data frame. 
If you look at this example transcriptome, you will see that the 3 genes have very different ranges of value. 
The ranges are orders of magnitudes different.  
Gene 1 ranges from 1 to 10.  
Gene 2 ranges from 100 to 1000.  
Gene 3 ranges from 10 to 100.   
You will also see that despite very different orders of magnitude, both gene 1 and 2 increases throughout the experiment,  
whereas gene 3 decreases. 

You can capture trends across different scales of magnitude using z-score. 
z score can be understood as how many standard deviations away from the mean for any given value.  
You can think of it as the differences in absolute magnitude are stabilized by their respective mean.  

The formula for z-score is:
z = (x - mean(x)) / sd(x). 

To calculate the z-score, we can easily make a new column using `mutate()`.  
But we need to remember that we want the z-score calculated for each gene separately.  
This is where `group_by()` comes in handy. 

```{r}
example_transcriptome_long_z <- example_transcriptome_long %>% 
  group_by(gene) %>% 
  mutate(z_score = (tpm - mean(tpm))/sd(tpm)) %>% 
  ungroup()

example_transcriptome_long_z
```

You can see that `group_by()` does not modify the data frame.  
But it modified how `mutate()` functions.  
We called `group_by(gene)`, which tells R to calculate the z-score for each gene. 
You can see that for genes 1 and 2, the z-scores increase from ~-1 to 1 through time points. 
It has the opposite trend For gene 3.  

The `group_by()` function saves you from having to do a for loop,  
where you will have to do it for each gene in the data frame, calculate means and sd, then calculate z score. 
With `group_by()`, the entire process is automated.  
Don't forget to `ungroup()` when you are done.  

# Summary statistics 
Tidyverse has an easy syntax called `summarise()` to calculate summary statistics such as mean and median from data frames. 
Summary statistics that can be plugged into `summarise()` includes:

* average: `mean()`
* median: `median()`
* stand deviation: `sd()` 
* sample size: `n()` 

And many others. 

Let's try it out on an hypothetical experiment. 
```{r}
# You can ignore this chunk. 
trt <- rnorm(n = 10, mean = 10, sd = 1)
ctrl <- rnorm(n = 10, mean = 6, sd = 1)
location <- c(rep("A", 5), rep("B", 5))

example_experiment <- data.frame(
  ctrl,
  trt,
  location
) %>% 
  pivot_longer(names_to = "treatment", values_to = "response", cols = 1:2)

example_experiment
```

In this experiment, we have two treatments, control (ctrl) and treated (trt).  
The experiment was repeated at two locations, A or B.  
At each location, 10 samples were collected. 
The dependent variable is recorded in the "response" column. 

Let's say I want to find the mean, median, sd, and n for the whole experiment, 
we will do: 
```{r}
example_experiment_s <- example_experiment %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    sd = sd(response),
    n = n()
  ) 

example_experiment_s
```

The resulting data frame has 1 row and 4 columns, with each column being a summary statistic. 

## Summary statistics by group. 
What is more informative will be to calculate summary statistics by group (ctrl vs. trt)  
so that we can begin to understand if the treatment had an effect on the dependent variable. 
To do that, we simply need to perform `group_by(treatment)` before running `summarise()`.
```{r}
example_experiment_s_by_group <- example_experiment %>% 
  group_by(treatment) %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    sd = sd(response),
    n = n()
  ) %>% 
  ungroup()

example_experiment_s_by_group
```

1.  `group_by()` selects which columns you want to summarize from.
    In this case you just want each group to be a group.
2.  `summarise()` allows you to call the statistical functions, such as mean, sd, and n.
3.  It's a good practice to `ungroup()` after you are done.
    Most of the time it doesn't matter, but it will create problems if you don't and you try to switch/modify grouping later.

Now you just get 2 rows and 4 columns with each row as a group. 

Let's say I want summary statistics for each group and for each location. Can I do that?
Yes, you just need to include location in `group_by()`. 
```{r}
example_experiment_s_by_group_location <- example_experiment %>% 
  group_by(treatment, location) %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    sd = sd(response),
    n = n()
  ) %>% 
  ungroup()

example_experiment_s_by_group_location
```

1.  This time you need to select treatment and location in `group_by()`
    because you want the mean for each treatment and each location.
2.  In `summarise()` you call the mean, median, sd, and n functions.
3.  It's a good practice to `ungroup()` after you are done.

Now each location by treatment combination is a row. 
Because we have 2 treatments and 2 locations, we should have 2*2 = 4 rows total.  
In this experiment, we can see that ctrl vs. trt seemed quite different for both locations. 

I hope you can see that `group_by()` followed by `summarise()` is very convenient and powerful. 
It saves you from having to use for loops to calculate statistics for each group. 
Instead, `group_by()` automates that. 

## Summary statistics by min and max. 
Let's go back to our hypothetical transcriptome.
```{r}
example_transcriptome_long
```

Say we want to find the sample where each gene is most highly or most lowly expressed in. 
To do that, we will need the `slice()` family functions. 

```{r}
example_transcriptome_highest_exp <- example_transcriptome_long %>% 
  group_by(gene) %>% 
  slice_max(order_by = tpm, n = 1) %>% 
  ungroup()

example_transcriptome_highest_exp
```
It functions as both summary and filtering. 

* First, it finds the min or max, depending on if you called `slice_min()` or `slice_max()`. 
* Then, it filters the row that has the min or max value. 

The resulting table has each gene as a row, and only contains the row with the highest expression. 
Since we called `n = 1` in `slice_max()`, it only returns the top expressing time point. 
If we used `slice_min()` here, we will find the lowest expressing time point instead. 

Let's try another example. 
Say we want to find top 3 highest income country per continent in the year 2010, using `income_tidy_continent`. 
```{r}
income_top3_by_continent_2010 <- income_tidy_continent %>% 
  filter(year == 2010) %>% 
  group_by(continent) %>% 
  slice_max(order_by = income, n = 3) %>% 
  ungroup()

income_top3_by_continent_2010
```

1. We first filter for `year == 2010`. 
2. Then we group by continent. 
3. Then we call `slice_max()` with `n = 3` to extract the top 3. 
4. Lastly, we `ungroup()` when we are done. 

# In class activity 
## Q1 
1. Load "S7_activity1.csv" and "S7_activity2.csv" from the Data folder. 
These are two datasets downloaded from the [Gapminder foundation](https://www.gapminder.org/data/). 

* "S7_activity1.csv" contains child mortality (0 - 5 year old) per 1000 born.
* "S7_activity2.csv" contains average births per family.    

These were recorded/estimated from year 1800 and projected all the way to 2100.  
Join these two data frames together after tidying them. 
Also join the income data frame (in tidy form) to make a larger table that contains mortality, births, income, and continent columns. 
Hint: you will need to use `by = c("country", "year")` inside `inner_join()` to specify that you need to match both the country and year columns. 

## Q2
Filter the data for just Asia. Find the top 10 countries based on income between year 1945 and 2015.   
Hint: use `group_by()` to automate calculations across years. 

## Q3 
Filter the data for year 2010. 
Classify countries based on the birth rate. 
If the birth rate is > 2, classify as "Class 1". 
Otherwise, classify as "Class 2". 
How do average income and mortality compare between these 2 classifications? 
Hint: you will need to call `mean()` twice in `summarise()`, once for income and once for mortality. 





