---
title: "Session 8 in-class activity"
author: "Chenxin Li"
date: "2026-06-17"
output: html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, we will cover: 

1. How are data represented in visualizations? 
2. What is "the grammar of graphics"? 
3. Introduction to ggplot: mapping variables to axis and aesthetics. 

# Packages
`tidyverse` will be doing all the heavy lifting. 
Please install `RColorBrewer` using the "Packages" panel at the lower right. 
We will be using some color sets from `RColorBrewer`. 
```{r}
library(tidyverse)
library(readxl)
library(RColorBrewer)
```

# How are data represented in visualization? 
As an example, let's simulate some data: 
```{r}
set.seed(666)
Example_data1 <- rbind(
  rnorm(3, mean = 5, sd = 0.5) %>%
    as.data.frame() %>% 
    mutate(TP = 1),
  rnorm(3, mean = 8, sd = 0.5) %>% 
    as.data.frame() %>% 
    mutate(TP = 2),
  rnorm(3, mean = 10, sd = 0.5) %>%
    as.data.frame() %>% 
    mutate(TP = 3)
) %>% 
  rename(response_sec = ".")
```

| response | time |
|:--------:|:----:|
| 5.376656 |   1  |
| 6.007177 |   1  |
| 4.822433 |   1  |
| 9.014084 |   2  |
| 6.891563 |   2  |
| 8.379198 |   2  |
| 9.346907 |   3  |
| 9.598740 |   3  |
| 9.103880 |   3  |

As you can see, the data are stored in this table. This is already a tidy data frame. 
It has two columns, "response" and "time". It appears each time point has 3 observations. 

If you were to just print this table, it will be a perfectly correct representation of the data.  
For example, if you read the numbers, you probably notice the response variable goes up as time progresses.  

But that is NOT the point of data visualization. 
The purpose of data visualization is to create a visual and intuitive representation of the data. 
A good data visualization allows us to gain insight into the data at a glance. 
That being said, what are the ways to visually represent data? 

* By position, such that the data is represented by position along the axis.    
* In bar plots, values are represented by the distance from the x axis, and thus the length of the bar.   
* Data can be represented by color, such as in heat maps.   
* Lastly, data can be represented by how elements are arranged relative to each other, such as how rows & columns are grouped and ordered in a heat map. 

# ggplot: data visualization using the grammar of graphics
We will use `ggplot()` to produce graphs. 
The "gg" in `ggplot()` stands for grammar of graphics, a theoretical framework to visualize data. 
In grammar of graphics, each visualization has the following layers: 

1. Coordinates: what are on x and y axis? 
2. Geometric objects ("geom"): e.g., bars, points, lines... 
3. Mapping: mapping colors, size, and transparency to geometric objects.

For more resources, you can look at the [ggplot cheat sheet](https://statsandr.com/blog/files/ggplot2-cheatsheet.pdf). 

Let's build a scatter plot for our example data. 
Usually I start with the data table and pipe it into `ggplot()` using the pipe `%>%` syntax.
Inside the `ggplot()` function, there is this `aes()` argument. 
Inside the `aes()` argument, it takes the x and y variables. 

```{r}
Example_data1 %>% 
  ggplot(aes(x = TP, y = response_sec)) +
  theme_classic()
```

If we just do that, we get a blank graph with time on x axis and response on y axis. 
The `theme_classic()` only change the appearance of the background. Read more [here](https://ggplot2.tidyverse.org/reference/ggtheme.html).
ggplot comes with a collection of themes. My go-to is "classic". It has white background and axis lines, as you can see in this example. 

The graph is blank because we haven't added any geometric objects (geom) yet. 
For the sake of this example only, let's make a scatter plot. We will need some points. 
The command to add points in `ggplot` is `geom_poit()`. 

```{r}
Example_data1 %>% 
  ggplot(aes(x = TP, y = response_sec)) +
  geom_point() +
  theme_classic()
```
This is how the graph looks after we added points. 
To include more layers, you add (using the `+` syntax) layers to the existing code. 
After the `ggplot()` line which initiates the graph, I usually add the geom layers before the other layers, then add the `theme` layer last. 

It's very important to have legible axis labels. 
To change the axis, you will need `labs(x = "something", y = "something")` 

```{r}
Example_data1 %>% 
  ggplot(aes(x = TP, y = response_sec)) +
  geom_point() +
  labs(x = "Time point",
       y = "Response") +
  theme_classic()
```
For a minimal plot, we are done!  

# Mapping variables to ggplot geom 
The example data we are using only have 2 variables: time and response. What if we have more than 2? 
To demonstrate that, let's use another example. 

I am going to read into child mortality, birth rate, and income data.  
For the sake of this example, we will just use year 1945. 

First, read in data.  
Second, re-shape data into tidy format. 
```{r message=F}
income <- read_csv("../Data/S6_activity1.csv")
income_tidy <- income %>% 
  pivot_longer(cols = !country, names_to = "year", values_to = "income") 

mortality <- read_csv("../Data/S7_activity1.csv")
mortality_tidy <- mortality %>% 
  pivot_longer(cols = !country, names_to = "year", values_to = "mortality") 

births <- read_csv("../Data/S7_activity2.csv")
births_tidy <- births %>% 
  pivot_longer(cols = !country, names_to = "year", values_to = "births") 
```

Third, join them together. 
```{r}
income_mortality_births <- income_tidy %>% 
  inner_join(mortality_tidy, by = c("country", "year")) %>% 
  inner_join(births_tidy, by = c("country", "year"))

head(income_mortality_births)
```

Fourth, filter for year 1945. 
```{r}
income_mortality_births_1945 <- income_mortality_births %>% 
  filter(year == 1945)
```

Let's say we want to make a scatter plot with birth rate on the x axis and mortality on the y axis, and color the dots by income.
We can make a plain scatter plot first. 
We can also add a title by using the `title` option in `labs()`.  

```{r}
income_mortality_births_1945 %>% 
  ggplot(aes(x = births, y = mortality)) +
  geom_point() +
  labs(x = "Birth rate",
       y = "Child mortality/1000 born",
       title = "Year 1945") +
  theme_classic()
```
To color dots (or any geom) based on a variable in the data frame, you open an `aes()` argument inside the `geom()`.  
Let's say we want to color by income. 
```{r}
income_mortality_births_1945 %>% 
  ggplot(aes(x = births, y = mortality)) +
  geom_point(aes(color = log10(income))) +
  labs(x = "Birth rate",
       y = "Child mortality/1000 born",
       title = "Year 1945") +
  theme_classic()
```

Note: income is quite unevenly distributed, so I log-transformed it. 
Now we have dots colored by income (in log10 scale).
We can use a different set of colors from the `RColorBrewer` package if you do not like the default color scale in ggplot. 
You can look at all the available color scales in `RColorBrewer` [here](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html). 
Let's use something easier to see than the default ggplot blue. 

You can change the color scale using the `scale()` family functions. 
```{r}
income_mortality_births_1945 %>% 
  ggplot(aes(x = births, y = mortality)) +
  geom_point(aes(color = log10(income))) +
  scale_color_gradientn(colours = brewer.pal(9, "YlGnBu")) + 
  labs(x = "Birth rate",
       y = "Child mortality/1000 born",
       title = "Year 1945") +
  theme_classic()
```
To provide a custom color scale that produces a color gradient, use `scale_color_gradientn())`.
Here is a whole family of `scale` functions in ggplot, which allows you to control the scales for colors and axis.  
You can read more about scales on the [ggplot cheat sheet](https://statsandr.com/blog/files/ggplot2-cheatsheet.pdf). 
I quite like the yellow-green-blue color gradient from `RColorBrewer`, which is what I used here. 
As you can see, low income countries have a higher birth rate and also higher child mortality.   

# Line plots with ggplot 
Line plot is an effective way to communicate each line is one group and one sequence of data.   
When you need to make a line plot, you could be doing one of the following things:  

1. Connecting individual data points, when each (x,y) position has a unique value.  
2. Connecting averages, when each independent variable position has multiple data points. 

## Connecting individual data points 
Let's use the crop yield data as an example. 
```{r}
crop_yield <-  read_csv("../Data/S6_example1.csv", col_names = T)
crop_yield_long <- crop_yield %>% 
  pivot_longer(cols = !Crop, names_to = "Year", values_to = "Yield") %>% 
  filter(is.na(Yield) == F) %>% 
  mutate(Year = as.numeric(Year))

head(crop_yield_long)
```

I went ahead and loaded it, converted it to tidy form, removed missing yield values, and changed the year to a numeric variable.   
Now it's all good to go. 
Let's make a line plot to showcase how yield has changed over the years for each crops. 
To do that, we will need `aes(x = Year, y = Yield)` inside `ggplot()`. 
We will use `geom_line()`.  
Since we need one line per crop, the syntax is `geom_line(aes(group = Crop))`. 

```{r}
crop_yield_long %>% 
  ggplot(aes(x = Year, y = Yield)) +
  geom_line(aes(group = Crop, color = Crop)) +
  scale_color_manual(values = brewer.pal(n = 8, "Dark2")) +
  labs(x = "Year", y = "Crop yield") +
  theme_classic()
```
I also asked R to color the lines by crop. 
Then I set the color to "Dark2" of the RcolorBrewer collection.  
I also manually set the x and y axis titles.  

There is one problem here. The yield of each crop are in different units, so it wouldn't be correct to plot them on the same plot. 
Also, there are no data for canola until the mid 90s. 
One solution is to make an individual sub-plot per crop. The jargon is "facet". 
The syntax is `facet_wrap(~variable_of_interest)`, in this case it would be `~Crop`.

```{r}
crop_yield_long %>% 
  ggplot(aes(x = Year, y = Yield)) +
  facet_wrap(~Crop, scales = "free") +
  geom_line(aes(group = Crop, color = Crop)) +
  scale_color_manual(values = brewer.pal(n = 8, "Dark2")) +
  labs(x = "Year", y = "Crop yield") +
  theme_classic()
```
I also added `scales = "free"` in `facet_wrap()`
such that each sub-plot will have a different scale that is more appropriate for the data in the subplot. 
We will talk more about using facets in later lectures. 
In general, facets are very powerful and should be considered when applicable. 

## Connecting means 
In the crop example, for each crop, there is only one value per year. 
For many experiments, there will be multiple replicates per treatment. 
In that case, we don't actually want to connect the individual data points. 
We want to connect the average of the replicates instead. 

Let's do an example (hypothetical data). 
```{r}
small_RNA <- read_excel("../Data/S8_example1.xlsx")
head(small_RNA)
```

Small RNA are RNAs that are between 20 to 25-nt long. 
You profiled small RNA in 3 cell types (sperm, egg, and somatic cells) using high throughput sequencing. 
Each cell types have 3 replicates. 
You recorded the % of small RNA relative to total reads. 

First we will need to turn this wide format table into a tidy data frame. 
Then we will need to create a new column called "cell_type". 
```{r}
small_RNA_tidy <- small_RNA %>% 
  pivot_longer(cols = !length_classes, names_to = "sample", values_to = "smRNA_abundance") %>% 
  mutate(cell_type = case_when(
    str_detect(sample, "sperm") ~ "sperm",
    str_detect(sample, "egg") ~ "egg",
    str_detect(sample, "somatic") ~ "somatic"
  ))

head(small_RNA_tidy)
```

Now we can make a line plot where each line is a cell type with RNA length on the x-axis and abundance on the y-axis. 
`ggplot()` has built-in summary statistics utility. 
The syntax is `stat_summary()`.  
Inside `stat_summary()`, if you need a line, you will call `geom = "line"`. It also requires the function of summary statistics, 
so you will call `mean_se`, which stands for mean and standard error. 
Then the rest is similar to normal `geom_line()`, where inside `aes()` you will call `group = cell_type`. 
Similarly, you can also add standard errors of mean as error bars using `geom = "linerange"` in another line of `stat_summary()`. 

```{r}
small_RNA_tidy %>% 
  ggplot(aes(x = length_classes, y = smRNA_abundance)) +
  geom_point(aes(color = cell_type)) + 
  stat_summary(geom = "line", fun.data = mean_se, aes(group = cell_type, color = cell_type)) +
  stat_summary(geom = "linerange", fun.data  = mean_se, aes(color = cell_type)) +
  scale_color_manual(values = c("tomato1", "seagreen", "dodgerblue2")) +
  labs(x = "smRNA length (nt)",
       y = "smRNA abundance (% of reads)",
       color = "cell type") +
  theme_classic()
```
I color the dots and lines by cell type. 
Manually set the axis titles, legend title, as well as the color set. 

# Saving ggplot to disk 
We haven't saved any of our graphs. 
To save our graphs to disk, we can use `ggsave()`, which supports vector (.svg), PDF, and pixel (.png) formats.  
You will need to specify the dimension of the output using width and height options. 
You will also need to specify export directory and file name. 

## Saving a ggplot as an object first 
There are two ways to export a graph to disk. 
The first way is to save your graph as an object first. 
```{r}
smRNA_line_plot <- small_RNA_tidy %>% 
  ggplot(aes(x = length_classes, y = smRNA_abundance)) +
  geom_point(aes(color = cell_type)) + 
  stat_summary(geom = "line", fun.data = mean_se, aes(group = cell_type, color = cell_type)) +
  stat_summary(geom = "linerange", fun.data  = mean_se, aes(color = cell_type)) +
  scale_color_manual(values = c("tomato1", "seagreen", "dodgerblue2")) +
  labs(x = "smRNA length (nt)",
       y = "smRNA abundance (% of reads)",
       color = "cell type") +
  theme_classic()

ggsave(plot = smRNA_line_plot, "../Results/S8_example1.svg", height = 3, width = 3)
ggsave(plot = smRNA_line_plot, "../Results/S8_example1.png", height = 3, width = 3)
```

Now I have saved the small RNA line plot to the Results folder as both svg and png formats. 
The default unit for height and width is inches. 
The best plot size is big enough to accommodate all text and elements, but small enough to not take up extra space on your paper or poster. 
3x3 is a good starting point, and you might need to adjust the export dimension a few times to find the best dimension. 

## Not saving ggplot as an object 
When you call `ggsave()` without specifying which plot to save, it will default to the last graph that you plotted.  
If you want to use this approach, make sure you save the graph right after it is plotted. 
```{r}
crop_yield_long %>% 
  ggplot(aes(x = Year, y = Yield)) +
  facet_wrap(~Crop, scales = "free") +
  geom_line(aes(group = Crop, color = Crop)) +
  scale_color_manual(values = brewer.pal(n = 8, "Dark2")) +
  labs(x = "Year", y = "Crop yield") +
  theme_classic()

ggsave("../Results/S8_example2.svg", height = 4, width = 6.5)
ggsave("../Results/S8_example2.png", height = 4, width = 6.5)
```

# In class activity 
## Q1 
Using the income, birth rate, and mortality data, 
graph income (in log10 scale) on the x axis and child mortality on the y axis from year 2000 to year 2010.  
Make sure to have legible axis titles. 
Hint: you can use facet by year, or use color by year.  
Save your graph using `ggsave()`. 

## Q2 
Check the first 6 lines of the built-in `ToothGrowth` dataset.  
Experimental animals were supplemented with either "VC" or "OJ" (the `supp` column). 
The independent variable is dose of supplement. 
The dependent variable is tooth length (`len`). 
Each dose and supplement has multiple replicates. 

Make a line plot, one line per supplement. 
The line should connect the means of each dose. 
Add error bars representing standard error of the mean. 
Color data points, line, and error bars by treatment.  
Color the OJ treatment with "grey", and the VC treatment with "tomato". 
Make sure to have legible axis and legend titles. 
Hint: you will need to use `stat_summary()` twice. 

Save your graph using `ggsave()`.  



 


